---
title: "Animal Shelter Outcomes \n for Labradors in Austin Animal Center"
author: 
- "Demi Dang"
- "Rosalia Hernandez"
- "Yanjie Qi"
date: "11/13/2019"
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center}\LARGE\includegraphics[width=6cm]{UCSBseal.jpg}\\[\bigskipamount]}
- \posttitle{\end{center}}
abstract: |
  Although Labradors are widely viewed as one of the most popular breeds in the United States, they are also a popular breed found in animal shelters [https://www.cleartheshelters.com](https://www.cleartheshelters.com/10-Most-Adoptable-Dog-Breeds-in-America-375215521.html). Taking a look at the outcomes of over 6,000 Labradors in the Austin Animal Center (the largest "no-kill" animal shelter in the U.S.), we build a Cox Proportional Hazards model to see how neuter status affects Labrador adoption rates.   
output: pdf_document
documentclass: report

---
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(survival)
#install.packages("tidyverse")
#install.packages("ggplot2")
library(tidyverse)
library(ggplot2)
#install.packages("survminer")
library(survminer)
library(knitr)
#library(gridExtra)
```


```{r data_organizing, echo = FALSE}
aac_shelter_outcomes <- read.csv("aac_shelter_outcomes.csv", stringsAsFactors = FALSE)

# Subset animal shelter data to include only labradors
labradors <- grep("Labrador", aac_shelter_outcomes$breed)
labradors.df <- aac_shelter_outcomes[labradors,]

# Exclude labs that have NULL and unknown values for sex_upon_outcome
null <- grep("NULL", labradors.df$sex_upon_outcome)
labradors.df <- labradors.df[-c(null),]

unknown <- grep("Unknown", labradors.df$sex_upon_outcome)
labradors.df <- labradors.df[-c(unknown),]


mixed <- grep("/", labradors.df$color)

labs_mixed <- labradors.df[mixed,]
labs_single <- labradors.df[-mixed,]

labs_mixed$color <- "Mixed"
labs_single$color <- "Solid"

# Combine the color data sets
labradors.df <- rbind(labs_mixed, labs_single)
```


```{r time_variables, echo = FALSE}
labradors.df$date_diff <- as.Date(as.character(labradors.df$datetime, format="%Y/%m/%d")) - 
                          as.Date(as.character(labradors.df$date_of_birth, format="%Y/%m/%d"))

```


```{r event_vector, echo = FALSE}
labradors.df <- labradors.df %>% mutate(event = case_when(outcome_type == "Return to Owner" ~ 1,
                                                              outcome_type == "Adoption" ~ 1,
                                                              outcome_type == "Rto-Adopt" ~ 1,
                                                              TRUE ~ 0
                                                              ))
```


```{r sex_variable, echo = FALSE}
for (i in 1:nrow(labradors.df)){
labradors.df$sex[i] <- tail(strsplit(as.character(labradors.df$sex_upon_outcome[i]), split = " ")[[1]], 1)
}
```


```{r neuter_variable, echo = FALSE}
for (i in 1:nrow(labradors.df)){
  a <- head(strsplit(as.character(labradors.df$sex_upon_outcome[i]), split = " ")[[1]], 1)
  if (a == "Intact")
    labradors.df$neuter[i] <- "Intact"
  else
    labradors.df$neuter[i] <- "Fixed"
}
```


```{r final_dataframe, echo = FALSE}
labradors.df <- labradors.df[,c("color", "event", "date_diff", "sex", "neuter")]

# Excluding labs that have negative date_diff because this could be due to input error: 844, 5658
labradors.df <- labradors.df[-c(844, 5658),]
```

## Data Source

The data we used was from the Austin Animal Center, downloaded from kaggle at [https://www.kaggle.com/aaronschlegel/austin-animal-center-shelter-outcomes-and](https://www.kaggle.com/aaronschlegel/austin-animal-center-shelter-outcomes-and). The raw data included cats, dogs, birds, livestock and other. Since we were only interested in labradors we created a subset of the data which included 2,128 Labrador and Labrador Mixed breeds for a total of 6,980 observations. We decided to delete one observation which had NULL values for `outcome` and `age upon outcome` since we thought our analysis without this observation would not change. We also excluded 48 data points where the gender and neuter status of the Labrador were not known. We split the column of `sex upon outcome` into two columns for analysis: `sex` with only Male/Female entries and `neuter` with Intact/Fixed entries where "Fixed" means the dog is either neutered or spayed and intact means it was not. Another covariate we changed was `color`. There were a ridiculous amount of "color" descriptions. Our favorite color that we saw was "peach." We decided to change this column to entries of either Mixed fur pattern or Solid fur patterns to denote dogs with more than one color versus dogs that are solid colored. Our final dataset included the following:

- `color`: The descriptive fur pattern of Mixed color fur pattern or solid color fur pattern.
- `event`: The event was 1 if the Labrador had been adopted, and censored was 0 if the Labrador had been transferred, missing (off-site, in foster care or in kennel), possible theft, euthanasia or death. 
- `date_diff`: The amount of time in days the Labrador spent in the shelter.
- `sex`: The gender of the dog.
- `neuter`: If the dog is fixed or intact.

```{r table1_datahead, echo = FALSE}
kable(head(labradors.df), caption = "Final Dataset")
```

## Research Question

On average how much time do Labradors spend in animal shelters before they go to a home?

We will plot a Kaplan-Meier curve and find the median time that Labradors spent in this animal shelter.

Does the color, gender or neuter status of a Labrador affect it's chance for adoption? 

We’ll build a Cox Proportional Hazards (Cox PH) model with the covariates neuter status, sex, and color to answer this question. We will also explore an Accelerated Failure Time (AFT) model as our extension component, to look at a comparison of survival times.


## Data Exploration

```{r datacountsummary, include=FALSE}
summary(as.factor(labradors.df$sex))
summary(as.factor(labradors.df$color))
summary(as.factor(labradors.df$neuter))
summary(as.factor(labradors.df$event))
```

Before we start our analysis we must understand our data. We found that we had a total of 6,929 observations. Out of these, 3211 are Female (46\%), 3713 have mixed fur pattern (53\%), and 5726 were fixed (82%). Using the quantile function on the length of time spent in the shelter, we found that 50\% of the dogs spent more than 395 days in the shelter, this is over one year! 

```{r densityquantile, include=FALSE}
quantile(labradors.df$date_diff, probs = c(0.25, 0.5, 0.75))
```

```{r densitydaysdistribution, echo = FALSE, fig.cap = "As a preliminary assesment of our data, and using the quantile function we find that half of the Labradors spend more than a year in the animal shelter (395 days, which is shown as the blue line). However, this form includes both censored and uncensored labradors which underestimates our value so we should take this figure lightly.", out.width="75%", fig.align="center"}
labradors.df$date_diff <- as.integer(labradors.df$date_diff)
ggplot(labradors.df) +
 aes(x = date_diff) +
 geom_density(adjust = 0.9, fill = "#fcbba1") +
 labs(x = "Days", 
      y = "Density", 
      title = "Density of Days in Shelter", 
      subtitle = "for Labradors in the Austin Animal Center") +
 theme_bw()+
geom_vline(xintercept = 396, linetype="solid", 
                color = "blue", size=0.5)

```


We plotted a Kaplan-Meier Curve without controlling for any variables. We noticed that the confidence interval curves were tight around the actual survival curve. This proves that the data is very detailed and the information is concrete because we have enough observations to do analysis on Labradors in the Austin Animal Center. 

```{r KM_all_labs, echo=FALSE, fig.align="center", out.width="75%", fig.cap="Kaplan-Meier curve"}
labradors.surv <- Surv(labradors.df$date_diff, labradors.df$event)
labradors.fit <- survfit(labradors.surv~1)
plot(labradors.fit,
     xlab = "Survival Time in Days",
     ylab = "Estimated Survival Probability",
     col = c("blue", "red", "red"),
     main = "Kaplan-Meier Curve for \n All Austin Animal Center Labradors")
abline(h = 0.5, col = "red")
```

```{r Median_KM, echo = FALSE}
labradors.fit
```

## Kaplan-Meier 

To visually see the survival curves we must plot Kaplan-Meier Curves for each covariate `color`, `sex` and `neuter` on time to adoption for labradors.
The Kaplan-Meier curves for sex in Figure 3 shows that male Labs have higher estimated survival rates since the blue male line is above the red female line. As time increases, male and female Labs have the same or similar survival rates because the lines look like they begin to overlap after the 3,000 days mark. 
The Kaplan-Meier curves for neuter status in Figure 4 shows that intact Labs (labs that have not been neutered) have higher estimated survival rates than fixed labs. This means that more fixed labs are being adopted or going to homes (having the event) than labs that have not been neutered. 
The Kaplan-Meier curves for fur patterns in Figure 5 show labs with single colored fur patterns have higher survival rates than those with mixed fur patterns. The gaps immediately widen after the 50% survival rate mark and begin to close again nearer to the bottom.


```{r KM_male_female, echo = FALSE, fig.cap="There is not a noticeable difference between the two curves so our pre-analysis hypothesis is that there is not going to be a significant difference.", out.width="75%", fig.align="center"}
labradors.fit2 <- survfit(labradors.surv ~ labradors.df$sex) 
plot(labradors.fit2,
     xlab = "Survival Time in Days",
     ylab = "Estimated Survival Probability",
     col = c("red", "blue"),
     main = "Kaplan-Meier Curve for \n Austin Animal Center Labradors by Sex")
legend("topright", legend = c("Female", "Male"), col = c("red", "blue"), pch=rep(19,2))
abline(h = 0.5, col = "red")
```

```{r KM_fixed, echo = FALSE, fig.cap="We see a bigger difference in the two survival curves between fixed and intact Labradors, mainly that fixed has a lower survival probability over time. However, this might be because there are many more observations on fixed labs than on intact labs, (fixed has about 4,000 more observations) so we will proceed with caution in our analysis because of this fact.", out.width="75%", fig.align="center"}
labradors.fit3 <- survfit(labradors.surv ~ labradors.df$neuter) 
plot(labradors.fit3,
     xlab = "Survival Time in Days",
     ylab = "Estimated Survival Probability",
     col = c("red", "blue"),
     main = "Kaplan-Meier Curve for \n Austin Animal Center Labradors by Fixed Status")
legend("topright", legend = c("Fixed", "Intact"), col = c("red", "blue"), pch=rep(19,2))
abline(h = 0.5, col = "red")
```


```{r KM_color, echo = FALSE, fig.cap = "There appears to be an even amount of labs with mixed and solid fur patterns. There seems to be a slight difference in survival probability between the two fur patterns with mixed having a lower survival probability.",out.width="75%", fig.align="center"}
labradors.fit4 <- survfit(labradors.surv ~ labradors.df$color) 
plot(labradors.fit4,
     xlab = "Survival Time in Days",
     ylab = "Estimated Survival Probability",
     col = c("red", "blue"),
     main = "Kaplan-Meier Curve for \n Austin Animal Center Labradors by Fur Pattern")
legend("topright", legend = c("Mixed", "Single"), col = c("red", "blue"), pch=rep(19,2))
abline(h = 0.5, col = "red")
```

## Log Rank Tests

The log-rank test allows us to compare 2 or more survival curves, more specifically, that there is a significant difference between the survival curves. The null hypothesis that we are testing is that there is no difference in survival curves. The low p-value means we reject our null hypothesis.


```{r logrank_sex, echo = FALSE}
survdiff(labradors.surv ~ labradors.df$sex)
```

```{r logrank_fixedstatus, echo = FALSE}
survdiff(labradors.surv ~ labradors.df$neuter)
```

```{r logrank_furpattern, echo = FALSE}
survdiff(labradors.surv ~ labradors.df$color)
```

After performing log-rank tests for the 3 covariates (`sex`, fur pattern, fixed status) individually, we see that the p-value for each log-rank test is less than $\alpha = 0.05$. The null hypothesis for each test is that the survival curves for the covariate are not significantly different, but since the p-values for each test are less than $\alpha = 0.05$, we can conclude that each of the covariates' survival curves are statistically significantly different. 

## Model Building

To pick the right covariates for our model we use the anova function on the full model which includes all three covariates that we are examining. The anova function below indicates that each of the covariates are significant to include in our model since each one has a p-value of less than 0.05. Since the last covariate also has a significant p-value then we can conclude that each previous covariate in the model is also included and thus also significant. In anova, order of covariates matter and each new covariate is added to the previous model and evaluated. 

```{r variableselection_anova, echo = FALSE}
fit1 <- coxph(Surv(date_diff, event) ~ neuter + color + sex, data = labradors.df)
anova(fit1)
```

```{r variableselection_forward, echo=FALSE, eval=FALSE}
# Comparing color to reduced model
reduced <- coxph(Surv(date_diff, event) ~ neuter, data = labradors.df)
full <- coxph(Surv(date_diff, event) ~ neuter + color, data = labradors.df)

lrt <- 2*(full$loglik[2] - reduced$loglik[2])

pchisq(lrt, df = 1, lower.tail=FALSE)

# Comparing sex to reduced model
reduced <- coxph(Surv(date_diff, event) ~ neuter + color, data = labradors.df)
full <- coxph(Surv(date_diff, event) ~ neuter + color + sex, data = labradors.df)

lrt <- 2*(full$loglik[2] - reduced$loglik[2])

pchisq(lrt, df = 1, lower.tail=FALSE)
```

 
We also use the backward elimination method in the step function which also indicates that we should include all three of our covariates. The step function uses the Akaike Information Criterion (AIC) to measure each model quality and gives us the best model for the specificied stepwise search (i.e. backward, forward, or both).

```{r variableselection_backwardelimination, echo = FALSE}
step(fit1, direction="backward")
```



## Model Checking

In order to build our Cox PH model we must check to see if the proportional hazard assumptions are met. We will do this by graphical evaluation of the log-log plots and by doing a residuals test using the cox.zph() function.

When evaluating the log-log plots we want to make sure that the survival curves don’t cross and that the curves look proportional to one another through time. These criteria imply that the proportional hazards assumption is met. Figure 6 shows the log-log plot for the `sex` covariate. We see that the curves cross and overlap both at the beginning and end of the curves. The suggests that the PH assumption is violated. Figure 7 shows the log-log plot for the `neuter` covariate. We see that the survival curves cross in the beginning, but then seem to be proportional until the very end. Because of this, we will also assume the PH assumption is violated. In Figure 8, the log-log plot for the `color` covariate shows an increase in the gap between the curves, and then a decline in the size of the gap through time. This suggests that the hazard proportion is not constant and that the PH assumtions do not hold.


```{r, warning=FALSE, out.width="75%", fig.align="center", fig.cap="The log-log plot for the sex covariate violates the PH assumptions we are checking because the curves appear to be crossing in various places.", echo = FALSE}
labs.fit1 <- survfit(Surv(date_diff,event) ~ sex,data = labradors.df)
plot(labs.fit1, 
     fun = "cloglog", 
     xlab = "Days",
     ylab = "-log(log(x))",
     main = "Log-log plot of Sex Covariate",
     col = 1:2,
     xlim=c(50,6500))
legend("bottomright",legend = c("male","female"),pch = rep(15,4),col = 1:2)
```

```{r, warning=FALSE, echo = FALSE, out.width="75%", fig.align="center", fig.cap="The log-log plot for the neuter covariate appears to be proportional except for the beginning of the curves where they cross a little and it doesn't appear to have constant hazard ratios through time since we see the the gap gets skinnier at the end."}
labs.fit2 <- survfit(Surv(date_diff,event) ~ neuter,data = labradors.df)
plot(labs.fit2, 
     fun = "cloglog", 
     xlab = "Days",
     ylab = "-log(log(x))",
     main = "Log-log plot for Neuter Covariate",
     col = 1:2,
     xlim=c(50,6500))
legend("bottomright",legend = c("Intact","Neutered"),pch = rep(15,4),col = 1:2)
```


```{r, warning=FALSE, echo = FALSE, out.width="75%", fig.align="center", fig.cap="The survival curves appear to get closer to each other as time progresses. This means that the hazard ratio is no longer constant and it depends on time."}
labs.fit3 <- survfit(Surv(date_diff,event) ~ color,data = labradors.df)
plot(labs.fit3, 
     fun = "cloglog", 
     xlab = "Time",
     ylab = "-log(log(x))",
     main = "Log-log plot for Color Covariate",
     col = 1:2,
     xlim=c(50,6500))
legend("bottomright",legend = c("Mixed","Single"),pch = rep(15,4),col = 1:2)
```


Using the Goodness-of-Fit test (table below) to further evaluate the PH assumptions, we test the null hypothesis that there is no correlation between the residuals and survival times. To do this we use the cox.zph() function in `R`. In the table below we see that for `color` and `sex` we reject the null hypothesis. This means that `color` and `sex` violate the PH assumptions. Notice that we do not reject the null hypothesis for the `neuter` variable which means that PH assumptions hold. Since this is a test and relied more on data than on visual analysis, we will follow the goodness-of-fit test conclusion that the PH assumptions hold for the `neuter` covariate. Our next step is to deal with the covariates that violate the PH assumption. We will do this by stratifying those covariates. 

```{r ph_assumptions_coxzph, echo = FALSE}
cox.zph(fit1, global=FALSE)
```



## Stratified Cox PH Model
```{r stratified_coxphmodel, echo = FALSE}
labs.fitSC = coxph(Surv(date_diff, event) ~ neuter + strata(color) + strata(sex), data = labradors.df)
summary(labs.fitSC)
```


## Interaction Terms

We use the anova function 
```{r interaction_sexcolor, echo = FALSE}
labs_interaction1 <- coxph(Surv(date_diff, event) ~ neuter + strata(sex)*strata(color), data = labradors.df)


anova(labs.fitSC, labs_interaction1)
```


```{r interaction_sexneuter, echo = FALSE}
labs_interaction2 <- coxph(Surv(date_diff, event) ~ strata(color) + strata(sex)*neuter, data = labradors.df)

anova(labs.fitSC, labs_interaction2)
```



```{r interaction_colorneuter, echo = FALSE}
labs_interaction3 <- coxph(Surv(date_diff, event) ~ strata(sex) + strata(color)*neuter, data = labradors.df)

anova(labs.fitSC, labs_interaction3)

```


```{r stratified_model_final, echo = FALSE}
labs.final <- coxph(Surv(date_diff, event) ~ neuter * strata(sex) + strata(color), data = labradors.df)

anova(labs.final)
```


```{r hazards_confidence, echo = FALSE}
ggforest(coxph(Surv(date_diff, event) ~ neuter, data = labradors.df), fontsize=0.7)
```


```{r baelinehazards, echo = FALSE}
bhaz = basehaz(labs.fitSC)

ggplot(bhaz) +
  geom_line(aes(x = time, y = hazard, colour = strata), size = 2)
```

## Extension: Accelerated Failure Times 

```{r lognormal, echo = FALSE}
#neuterfit <- survreg(Surv(date_diff, event) ~ neuter, data = labradors.df, dist = "lognormal")
#sexfit <- survreg(Surv(date_diff, event) ~ sex, data = labradors.df, dist = "lognormal")
#colorfit <- survreg(Surv(date_diff, event) ~ color, data = labradors.df, dist = "lognormal")

#summary(neuterwfit)
#summary(sexwfit)
#summary(colorwfit)
```

```{r survivalodds_neuter, echo = FALSE}
ps <- seq(0.01, 0.99, by = 0.01)
plot(log(labradors.fit3$time), 
     log(labradors.fit3$surv/(1 - labradors.fit3$surv)),
     pch = 18, type = "s", xlab = "Log Time",
     ylab = "Log Odds of Survival", col = c(2, 4),
     xlim = c(3, 9))
```


```{r weibull, echo = FALSE}
neuterwfit <- survreg(Surv(date_diff, event) ~ neuter, data = labradors.df, dist = "weibull")
sexwfit <- survreg(Surv(date_diff, event) ~ sex, data = labradors.df, dist = "weibull")
colorwfit <- survreg(Surv(date_diff, event) ~ color, data = labradors.df, dist = "weibull")

anova(neuterwfit)
anova(sexwfit)
anova(colorwfit)
```

 
```{r exponential, echo = FALSE}
neuterfit.e <- survreg(Surv(date_diff, event) ~ neuter, data = labradors.df, dist = "exponential")
sexfit.e <- survreg(Surv(date_diff, event) ~ sex, data = labradors.df, dist = "exponential")
colorfit.e <- survreg(Surv(date_diff, event) ~ color, data = labradors.df, dist = "exponential")

anova(neuterfit.e, neuterwfit)
anova(sexfit.e, sexwfit)
anova(colorfit.e, colorwfit)
```


```{r loglogistic, echo = FALSE}
neuterfit.ll <- survreg(Surv(date_diff, event) ~ neuter, data = labradors.df, dist = "loglogistic")
sexfit.ll <- survreg(Surv(date_diff, event) ~ sex, data = labradors.df, dist = "loglogistic")
colorfit.ll <- survreg(Surv(date_diff, event) ~ color, data = labradors.df, dist = "loglogistic")

summary(neuterfit.ll)
summary(sexfit.ll)
summary(colorfit.ll)
```


