---
title: "PSTAT 175 LAB B"
author: "Yanjie Qi"
date: "2019/10/20"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(survival)
library(tidyverse)
```

Import the datasets:
```{r}
vets <- read.table("~/Desktop/Fall 2019/PSTAT 175/LAB B/vets.txt", quote="\"", comment.char="")
heroin <- read.table("~/Desktop/Fall 2019/PSTAT 175/LAB B/heroin.txt", quote="\"", comment.char="")
```

## 1
```{r}
# create vectors
vet.time <- vets$V1
vet.cns <- vets$V2
vet.Surv <- Surv(vet.time, vet.cns)
vet.fit <- survfit(vet.Surv ~ 1, conf.int=0.95)
```
## a)
```{r}
# plot the kaplan-meier estimator with 95% CI
plot(vet.fit,main="Kaplan-Meier Curves \n for VA Medical Clinic Subjects",
     xlab="Time (until death) \n (in days, weeks, etc.)",
     ylab=expression(hat(S)(t)),col=c("steelblue","black","black"),lwd=2,
mark.time = TRUE,mark=18)
```

## b)
```{r}
## calculate estimate of the 75th, 50th, and 25th percentiles
x<-min(vet.fit$time[vet.fit$surv < .75])
y<-min(vet.fit$time[vet.fit$surv < .50])
z<-min(vet.fit$time[vet.fit$surv < .25])
print(c("75th percentile =", x, "50th percentile =", y, "25th percentile =", z))
```

## 2
load the dataset
```{r}
data(lung, package = "survival")
attach(lung)
```

## a)
create the vectors:
```{r}
lung.time <- lung$time
lung.cts <- lung$status
lung.surv <- Surv(lung.time, lung.cts)
lung.fit <- survfit(lung.surv ~ 1, conf.int = 0.95)
```
Plot the Kaplan-Meier estimator:
```{r}
plot(vet.fit,main="Kaplan-Meier Curves \n for lung cancer patients",
     xlab="Time (until death) \n (in days, weeks, etc.)",
     ylab=expression(hat(S)(t)),col=c("steelblue","black","black"),lwd=2,
mark.time = TRUE,mark=18)
```

## b)
find the time near 150 days:
```{r}
t<-max(lung.fit$time[lung.fit$time<150])
t
```
Get the estimate and confidence interval:
```{r}
# by using summary function
summary(lung.fit, times=t)
```

## c)
Find the median survival time:
```{r}
tm <- median(lung.fit$time)
tm
```
Get the estimate and confidence interval:
```{r}
summary(lung.fit, times=tm)
```

## d)
```{r}
# Set Up the Vector
lu.s = survfit(lung.surv~sex,data=lung)
lu.s
```
Plot the seperated estimators:
```{r}
plot(lu.s,xlab="Survival time",ylab = expression(hat(S)(t)),lwd=2,
     col=c("steelblue","magenta"), mark.time = TRUE,mark=18)
legend("topright",legend=c("Male","Female"),
       col=c("steelblue","magenta"),pch=rep(19,2))
```
Generally Speaking, women have better survival rates. It is consistent except in the time around 750, the survival rate is almost the same.

## e)
Get the estimate and confidence interval:
```{r}
summary(lu.s, times = tm)
```
From these intervals, we could conclude that female survival rate is still higher than male's, since, in the end, female have 0.4937 survival rate with CI (0.4152,0.5870), but as for male, they have 0.7045 survival rate with CI (0.6129,0.8098). Accordingly, female sample indeed have higher survival rate. It might not be the whole story, since the plot in part(d) tells us there is a certain time, male and female samples have the identical survival rate, so the female survival rate is not always higher than male survival rate.

## 3
## a)
```{r}
# create the vector needed
her.surv <- Surv(heroin$Time, heroin$Status)
her.fit <- survfit(her.surv ~ 1, conf.int = 0.95)
```
Plot the Kaplan-Meier Estimate with 95% CI:
```{r}
plot(her.fit,main="Kaplan-Meier Curves \n for methadone treatment clinics patients",
     xlab="Time (until death) \n (in days, weeks, etc.)",
     ylab=expression(hat(S)(t)),col=c("steelblue","black","black"),lwd=2,
mark.time = TRUE,mark=18)
```

## b)
extract mj and nj from the survfit function output:
```{r}
mj = her.fit$n.event
nj = her.fit$n.risk
```
calculate mj/(nj∗(nj−mj)) and the cumulative sum:
```{r}
Vj = mj/nj/(nj-mj)
cVj = cumsum(Vj)
```
Then calculate the 95% CI:
```{r}
lowerCI = log(her.fit$surv) - 1.96*sqrt(cVj)
upperCI = log(her.fit$surv) + 1.96*sqrt(cVj)
```
Plot the estimate of S(t):
```{r}
par(mar=c(5,5,4,2))
plot(her.fit$time,log(her.fit$surv),lwd=2,type="l",ylim=c(-1.5,0),
xlab="Survival time",ylab=expression(log(hat(S)(t))), col="steelblue")
lines(her.fit$time,lowerCI,lty=2,col=1,lwd=2)
lines(her.fit$time,upperCI,lty=2,col=1,lwd=2)
```

## c)
Transform the plot from part b:
```{r}
lowerCI = her.fit$surv - 1.96*sqrt(cVj)
upperCI = her.fit$surv + 1.96*sqrt(cVj)
par(mar=c(5,5,4,2))
plot(her.fit$time,her.fit$surv,lwd=2,type="l",ylim=c(0,1),
xlab="Survival time",ylab=expression(hat(S)(t)), col="steelblue")
lines(her.fit$time,lowerCI,lty=2,col=1,lwd=2)
lines(her.fit$time,upperCI,lty=2,col=1,lwd=2)
```
We could see plot in part c has wider C.I. than the one in part, and the difference between C.I. of two side in part b seems more constant than the one in part c.

## d)
we could get for one year what percentage of patients are in the clinic from summary:
```{r}
summary(her.fit, times=365)
```
we should use two-sided hypothesis.
Hypothesis Test:
H0: S >= 50% vs. Ha: S < 50%
```{r}
t = max(her.fit$time[her.fit$time<365])
her.fit$surv[her.fit$time==t]
```
test statistics：
```{r}
z = (log(her.fit$surv[her.fit$time==t])-log(0.5))/her.fit$std.err[her.fit$time==t]
z
```
As shown above, the test statistic is greater than the critical value 1.96, so we reject H0 and conclude that less than 50% are being
discharged in one year.

Then, we coud compute the p-value:
```{r}
pnorm(-z,mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)
```
So, the p-value is 0.0002124677.
## e)
```{r}
q1 <- min(her.fit$time[her.fit$lower<0.3]) 
q2 <- max(her.fit$time[her.fit$upper>0.3]) 
c(q1,q2)
```
The time interval for 70th percentile is 661 826.
```{r}
summary(her.fit, times = c(q1,q2))
```
Then, for 80th percentile:
```{r}
n1 <- min(her.fit$time[her.fit$lower<0.2]) 
n2 <- max(her.fit$time[her.fit$upper>0.2]) 
summary(her.fit,times = c(n1,n2))
```
```{r}
c(n1,n2)
```
We know time=1076 is the end of dataset and it is censored. Therefore, we could not use this way to find the 80th percentile. 
Instead, we use the estimate for smaller s(t).